```{r setup, echo=F}
locale <- Sys.setlocale("LC_TIME", "C")
options(tinytex.clean=FALSE) # to keep .aux file so glossary can be extracted
knitr::opts_chunk$set(echo=F, fig.align='center')
library(kableExtra)

if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "latex") {
  kable <- function(data, caption='', col.names=c()) {
    knitr::kable(data, booktabs=TRUE, digits = 2, format='latex', escape=F, caption=caption, col.names=col.names) %>% 
      kable_styling(latex_options =c("striped"))
  }
} else {
  kable <- function(data, caption='', col.names=c()) {
    knitr::kable(data, booktabs = TRUE, digits = 2, escape=F, caption=caption, col.names=col.names, format='markdown') %>%
      kable_styling(bootstrap_options=c('striped'))
  }
}
```

---
title: "Thesis Title"
subtitle: "Thesis Subtitle"
date: "`r format(Sys.Date(), '%d. %B %Y')`"

cover-page: "cover.pdf"
title-page-logo: "uis-logo-en.pdf"
title-page-logo-text:
    - "Faculty of Science and Technology"
    - "Department of Electrical Engineering and Computer Science"
title-page-logo-width: 3in

author-heading: "Bachelor's Thesis in Computer Science by"
author:
  - name: "Elias Nodland"
    affiliation: 1

internal-supervisor-heading: "Supervised by"
internal-supervisor:
  - name: "Hein Meling"
    affiliation: 1

affiliation:
  1: "University of Stavanger"

lang: en

toc: true  # table of contents
lot: true  # list of tables
lol: true  # list of listings
lof: true  # list of figures
glossaries: true # list of terms
has-frontmatter: true

documentclass: book
papersize: a4
fontsize: 11pt
fontfamily: times
# mainfont: Georgia

fancyhdr:
    lhead: ""
    rhead: ""

geometry:
  - inner=1.5in
  - outer=1.0in
  - top=0.6in
  - bottom=0.8in

float:
  placement:
    figure: ht
    table: ht

headheight: 20pt
footheight: 14pt
headrulewidth: 0pt
footrulewidth: 0pt

bibliography: bibliography.bib
csl: ieee.csl
csl-refs: true
link-citations: true
---

```{r child='ch01-introduction/introduction.Rmd'}
```
\cleardoublepage

```{r child='ch02-background/background.Rmd'}
```
\cleardoublepage

```{r child='ch03-related-work/related-work.Rmd'}
```
\cleardoublepage

```{r child='ch04-architecture/architecture.Rmd'}
```
\cleardoublepage

```{r child='ch05-implementation/implementation.Rmd'}
```
\cleardoublepage

```{r child='ch06-discussion/discussion.Rmd'}
```
\cleardoublepage

```{r child='ch07-conclusion/conclusion.Rmd'}
```
\cleardoublepage

# References {-}

<div id="refs"></div>

\cleardoublepage

# (APPENDIX) Appendices {-}

```{r child='A-original-proposal/original-proposal.Rmd'}
```
