```{r setup, echo=F}
locale <- Sys.setlocale("LC_TIME", "C")
options(tinytex.clean=FALSE) # to keep .aux file so glossary can be extracted
knitr::opts_chunk$set(echo=F, fig.align='center', out.width='3.5in')
library(kableExtra)

if (knitr::opts_knit$get("rmarkdown.pandoc.to") == "latex") {
  kable <- function(data, caption='', col.names=c()) {
    knitr::kable(data, booktabs=TRUE, digits = 2, format='latex', escape=F, caption=caption, col.names=col.names) %>%
      kable_styling(latex_options =c("striped"))
  }
} else {
  kable <- function(data, caption='', col.names=c()) {
    knitr::kable(data, booktabs = TRUE, digits = 2, escape=F, caption=caption, col.names=col.names, format='markdown') %>%
      kable_styling(bootstrap_options=c('striped'))
  }
}

library(DiagrammeR)
```

---
title: "The Gotos framework"
subtitle: "for teaching operating systems to undergrads"
date: "`r format(Sys.Date(), '%d. %B %Y')`"

cover-page: "cover.pdf"
title-page-logo: "uis-logo-en.pdf"
title-page-logo-text:
    - "Faculty of Science and Technology"
    - "Department of Electrical Engineering and Computer Science"
title-page-logo-width: 3in

author-heading: "Bachelor's Thesis in Computer Science by"
author:
  - name: "Elias Nodland"
    affiliation: 1

internal-supervisor-heading: "Supervised by"
internal-supervisor:
  - name: "Hein Meling"
    affiliation: 1

affiliation:
  1: "University of Stavanger"

lang: en

glossaries: true # list of terms
has-frontmatter: true

documentclass: report
papersize: a4
fontsize: 11pt
fontfamily: times

fancyhdr:
    lhead: ""
    rhead: "\\rightmark"

geometry:
  - inner=1.5in
  - outer=1.5in
  - top=1.0in
  - bottom=1.0in

float:
  placement:
    figure: ht
    table: ht

headheight: 20pt
footheight: 14pt
headrulewidth: 0pt
footrulewidth: 0pt

biblatex: true
bibliography: bibliography.bib
biblio-style: ieee
csl: ieee.csl
csl-refs: true
link-citations: true
always_allow_html: true

# links are added as footnotes
header-includes:
  - \let\oldhref\href
  - \renewcommand{\href}[2]{#2\footnote{\url{#1}}}
---

\pagenumbering{roman}

```{r child='preface/abstract.Rmd'}
```

\cleardoublepage

```{r child='preface/acknowledgements.Rmd'}
```

\cleardoublepage

\raggedbottom

\tableofcontents
<!--
\listoffigures
\listoftables
\lstlistoflistings
\printglossaries
-->

\cleardoublepage

\pagenumbering{arabic}

# (PART) INTRODUCTION {-}

```{r child='01-introduction/include.Rmd'}
```

\cleardoublepage

```{r child='02-related-work/include.Rmd'}
```

# (PART) MAIN BODY {-}

\cleardoublepage

```{r child='03-architecture/include.Rmd'}
```

\cleardoublepage

```{r child='04-implementation/include.Rmd'}
```

\cleardoublepage

```{r child='05-assignments/include.Rmd'}
```

\cleardoublepage

# (PART) REFLECTION {-}

```{r child='06-discussion/include.Rmd'}
```

\cleardoublepage

```{r child='07-conclusion/include.Rmd'}
```

\cleardoublepage

# (APPENDIX) Appendices {-}

```{r child='A-original-proposal/include.Rmd'}
```

\cleardoublepage

```{r child='B-shootdown-alternative/include.Rmd'}
```

\cleardoublepage

```{r child='C-user-space/include.Rmd'}
```

\cleardoublepage

# References {-}

<div id="refs"></div>
