# Implementation

This chapter describes the implementation of the multiple systems and sub-systems found in the Gotos project.
Gotos requires Go version 1.18 or later.

Most of this chapter is dedicated to describing the RISC-V emulator, which is implemented in 3500 lines of code.

---

> *This somewhat large size is attributed to many comments and the repetitive implementation of one function for each instruction.*
> *The alternative is a large `switch`/`case` chain, but that would be sacrificing code readability.*

---

The code can be found on GitHub\ @github-gotos, or on GitLab\ @gitlab-gotos under our alias *mumblingdrunkard*.
It is an open source effort and is licenced under the traditional MIT license\ @licence-mit.

We give only a brief overview as we cannot possibly cover every function and interaction in the system.
Functions included here will usually leave out error/exception-handling and other "trivialities".
Still, this overview should give an adequate explanation of the core operation of the emulator.

For example uses, we refer to the draft included in appendix\ \@ref(ap:book).

```{r child='01-structure.Rmd'}
```

```{r child='02-main-loop.Rmd'}
```

```{r child='03-atomics.Rmd'}
```

```{r child='04-memory-controller.Rmd'}
```

```{r child='05-interrupt-sources.Rmd'}
```

## Implementation summary

In general we have followed a few guidelines to ensure performance stays good and the structure of the code is sound.

- return early,
- arrays over slices for critical core structures,
- panic on misuse, and
- semantic naming.

We have also written code with students in mind.
This means that the code should be resistant to misuse and error messages are clear and obvious when misuse leads to a panic.
