# Implementation

This chapter describes the implementation of the multiple systems and sub-systems found in the Gotos project.
Gotos requires Go version 1.18 or later.

Most of this chapter is dedicated to describing the RISC-V emulator, which is implemented in 3500 lines of code.
The code can be found on GitHub\ @github-gotos, or on GitLab\ @gitlab-gotos under our alias *mumblingdrunkard*.
**Note** that the code in these repositories is likely to be slightly behind the code presented in this chapter as the code here is developed on a secret branch until this thesis is graded.

---

> This somewhat large size is attributed to many comments and the repetitive implementation of one function for each instruction.
> The alternative is a large `switch`/`case` chain, but that would be sacrificing code readability.

---

```{r child='01-structure.Rmd'}
```

```{r child='02-main-loop.Rmd'}
```

```{r child='03-memory-controller.Rmd'}
```

```{r child='04-miscellaneous.Rmd'}
```

```{r child='05-examples.Rmd'}
```
