## Processor architecture - RISC-V in Gotos

The processor emulator included in Gotos serves as a basis for the framework to build on.
It is a trivial implementation of a RISC-V emulator, supporting only a 32 bit architecture (thus omitting a number of instructions, but keeping a decent target).

The main goal is to provide a feature-full processor that closely mimics the behaviour of real processors in single- or multicore systems, and with acceptable performance.
For these reasons, we omit multiple details of the RISC-V specification, opting for only those features that are relevant and required for OS development.

In this section we provide a description of the relevant parts of the RISC-V specification and where Gotos adheres or diverges from this, along with some justifications for the divergence.
We also describe briefly the different components that have been used to achieve the goal of efficient multicore execution.

### The RISC-V instruction set architecture (ISA) {#arch:isa}

Gotos uses a user-mode RISC-V emulator for processes running on the system.
This is similiar to Nachos' MIPS emulator\ @nachos-report.
This section contains a short description of the RISC-V instruction set architecture.

The RISC-V specification defines a processor with 31 general purpose registers and one special register that is hard-wired to all 0s.
It specifies a base instruction set -- RV32I -- of 40 instructions that operate on values in these registers.
This is the minimal instruction set that a RISC-V processor can reasonably support.

It also specifies several extensions that add capabilities such as floating-point arithmetic, or atomic operations.

#### The G extension

The G extension is a collective extension used to refer to the IMAFDZicsr Zifencei extensions all together.
Intended as a reasonable target for applications running on modern systems:

- I, the base RISC-V instruction set of 40 instructions, contains the bare minimum for many applications such as addition and subtraction, and control-flow instructions.
- M, extension for multiplication and division.
- A, extension for atomic operations.
- F, extension for single precision floating-point arithmetic (disabled by default).
- D, extension for double precision floating-point arithmetic (disabled by default).
- Zicsr, extension for user-mode control and status registers (\acrshort{csr}).
  Rarely used and probably should not be for user-mode programs as we are unsure about the correctness of its implementation (disabled by default).
- Zifencei, extension for instruction fence. Contains only one instruction that forces a core to flush its instruction cache.

The Gotos RISC-V emulator implements all of these extensions, allowing most system-independent code to compile and execute without any problems.
The F, and D extensions are non-compliant, but functional.
Applications should not expect fully compliant floats.

We expect students will not be writing many of their own applications as most applications used for testing the operating system should already be provided by the labs.
Still, the G extension provides a solid base to build on and it should be possible to use the emulator to teach basic assembly if desired.

However, test-applications should preferably be written only with the I, M, A, and Zifencei extensions which are enabled by default (and cannot be disabled).
The F, D, and Zicsr extensions require more work and quality-assurance before they can be reliably used in an application and are disabled by default.
If not enabled, instructions from these three extensions will raise a `TrapIllegalInstruction`.

For the Zicsr extension, note that it being disabled does not mean that the *Control and Status Register* file (*file* in this context having the computer-architecture definition\ @wikipedia-register-file) is disabled.
This register file is required for the OS to properly work.
The Zicsr is specifically about user-mode support for csr interaction.

See appendix\ \@ref(ap:user-space) for how to write and compile applications for the emulated processor, recommended compilation settings, and more.

#### Instruction encoding

Instructions are encoded in a structured manner in 32 bits\ [@rv5vol1, chapter 24].
Groups of similar instruction follow similar patterns.
Structure in some instructions is "sacrificed" for the sake of structure in others.
Examples of this are the branching instructions with *immediate offsets* (offset encoded directly into instruction, as opposed to being fetched from a register), one of which is shown in figure\ \@ref(fig:ifuckinghatethis).

```{r ifuckinghatethis, fig.cap='The encoding structure of a conditional branch instruction such as BEQ or BNE', out.width='100%'}
knitr::include_graphics(normalizePath('fig/cond-branch.pdf'))
```

All instruction have an **opcode** in the seven last bits of the 32-bit instruction.
This code does not fully define the instruction and is more a marker of instruction *type*.
The **opcode**s Gotos uses are: **op-imm**, **lui**, **auipc**, **op**, **jal**, **jalr**, **branch**, **load**, **store**, **misc-mem**, **system**, **amo**, **load-fp**, **store-fp**, **op-fp**, **fmadd**, **fmsub**, **fnmsub**, and **fnmadd**.

Some of these codes are single instructions (e.g. **lui** or **fnmsub**); others require more information (e.g. **op** or **system**).
For those instructions that require more information, they include a **funct3** or **funct7** -- and in some cases, even **funct12**\ [@rv5vol1, section 2.8] -- field which specifies the instruction further\ [@rv5vol1, section 2.2, 2.3].

### The RISC-V privileged architecture

The RISC-V specification also defines a privileged architecture.
This specification contains information about what a RISC-V implementation must provide for certain software like OSes to function.
Gotos heavily diverges from the specification at this level as the processor emulator is not intended to run full systems such as Xv6 or Linux.
Other projects already accomplish this task and do it with much higher efficiency and correctness.

The privileged specification still serves as inspiration for several mechanisms in Gotos.
One example of this is all of the privileged instructions as none of them are available to programs executing on a core.
Their effects are necessary for an OS, so they have to be achieved through other interfaces.

`sfence.vma` is a critical instruction that ensures a core can see updates made to the virtual memory configuration.
This instruction is not provided in the implementation and so its effect has to be achieved by other means.
Instead, Gotos `Core` provides `SFENCE_VMA()` which should be called on the core when performing an address space switch or something that otherwise requires a local TLB flush.
Note that this instruction is ***not*** available to user-space programs and there is no path of execution that will ever lead to it being executed by a user-space program.

#### Control and Status Registers (CSRs)

The privileged architecture is heavily dependant on special registers called *Control and Status Registers*.
There are 4096 of these registers, and their purposes are described in meticulous detail in the privileged specification\ [@rv5vol2, chapter 2].

We implement only a few of these registers:

- **fflags**, **frm**, and **fcsr** are dedicated to controlling IEEE floating-point behaviour on the system\ [@wikipedia-ieee754; @rv5vol2, section 11.2]; they are not used much, or at all (see section\ \@ref(arch:floating-point)),
- **mhartid** which contains the ID of the core (RISC-V uses the term *hart*, which is slightly more abstract than a core, but we will refer to them as cores to avoid confusion),
- **mepc**, **mcause**, and **mtval**, which are used for trap setup and handling, and
- **satp**, a register that controls virtual memory addressing and protection.

While we could avoid using CSRs completely, we implement them to stay as close as possible to the specification.

### Instruction fetch and execute

The emulator included in Gotos executes instructions in order and one at a time.
After checking interrupts, it both fetches and executes in the same cycle.
Instructions are guaranteed to complete in the order they are fetched.

Instruction fetch is no different from other reads from memory and goes through the same pipeline of translation and cache lookup.

### Floating-point numbers {#arch:floating-point}

The emulator in Gotos mostly follows the RISC-V specification and should give correct results for any user-space application that might be interesting in an OS course.
RISC-V provides IEEE-compliant floating-point arithmetic through the F and D extensions.
The behaviour of floats in RISC-V can be modified by writing configuration values to specific registers.

While we implement *some* version of floating-point numbers, we do not comply fully with the specification.

We imagine that any real use of floating-point numbers would be in performance critical applications such as in a real-time game.
We accept the compromise of a non-compliant implementation for the sake of improved performance.

### Atomic operations {#arch:atomics}

RISC-V uses load-reserved/store-conditional primitives for atomic operations\ [@rv5vol1, chapter 8].

#### A bit about load-reserved/store-conditional

RISC-V achieves atomicity accross multiple cores by using *reservation sets*.
A core can perform a `lr.w` instruction which will atomically load a word, and register a reservation that subsumes the loaded bytes.
Reservations in Gotos are tracked on a cache-line and use only physical addresses.

This reservation may be invalidated for a number of reasons such as:

- a store-conditional to the reserved region occured on a different core,
- another reservation was made by the same core, or
- a store is made to the reserved region by an external device.

At a later point the core may perform an `sc.w` instruction, which will attempt to store a word.
This store will only be successful if the reservation for that word is still valid.

Cores can not hold more than one valid reservation at a time, and reservations should be invalidated when a context switch occurs.

> "*A store-conditional instruction to a scratch word of memory should be used to forcibly invalidate any existing load reservation:*
>
> - *during a preemptive context switch, and*
> - *if necessary when changing virtual to physical address mappings, such as when migrating pages that might contain an active reservation.*"
>
> -- RISC-V Instruction Set Manual - Vol. I [@rv5vol1, page 50].

Load-reserved/store-conditional has benefits and disadvantages compared to something like *compare-and-swap* and can make debugging concurrent applications especially difficult as stepping through programs may inadvertently invalidate reservations so that none of the threads can make progress.

To prevent live-lock, RISC-V has architectural guarantees that a LR/SC-pair will eventually be successful as long as the SC instruction comes within 16 instructions after the LR instruction\ [@rv5vol1, section 8.3].

#### Reservation sets in Gotos

Gotos has `ReservationSets`, which is a shared struct that tracks reservations for all cores in the system.
Each core can have at most one reservation active at a time, and the shared struct must be locked when performing operations.

### Memory synchronisation

Besides atomics, RISC-V has instructions to provide certain guarantees of *observability*.
RISC-V has a *relaxed memory model*\ [@rv5vol1, page 27].
Where some architectures may require that different cores snoop on other cores' data/address-lines to synchronise caches, RISC-V uses special instructions to ensure that a write in one core is observable in other cores in the system.

These are called *fences* or *memory barriers* and they place ordering constraints on memory operations before and after the instruction, hence the name fence.
The specification does not place any restrictions or requirements for *how* these fences may be implemented other than that the rules of observability are followed\ [@rv5vol1, section 2.7].

Fences apply only to one core, meaning they cannot be used to achieve atomics.
They are useful when synchronisation is required, but the actual signalling of the synchronisation should be handled using atomics.
An example of this is a lock that uses `amoswap` (atomic swap) to acquire a mutex, then uses a `fence` to ensure the processor is reading the latest data (as changes may have been made by other cores that held the mutex).
A corresponding release function would call `fence` one last time to ensure data is written back to main memory after modifying data behind a mutex.

To prevent excessive flushing or invalidation, a fence may be modified to only order operations before or after the fence such that our lock/release becomes: (1) lock, invalidate cache; (2) only writeback, don't invalidate, then release mutex.

Gotos currently uses a conservative implementation and fully flushes and invalidates the cache on every `fence`.
We chose this approach as it is the safest and is guaranteed to be correct ^[Correct in the sense that, while performance may be impacted, execution is well-defined.].

### Memory protection, attributes, and virtualisation {#sec:mem-virt}

RISC-V provides several ways to set rules for memory access:

*Physical Memory Protection* (PMP) registers are special CSRs (\acrshort{csr}) that define rules for how physical may be accessed. This is an optional feature.

*Physical Memory Attributes* (PMA) registers are special CSRs (\acrshort{csr}) that define attributes of regions of memory such as:

- which atomic operations are permitted,
- reservation guarantees,
- access-alignment,
- memory-ordering,
- coherence and cacheability, and
- idempotency.

These regions are usually defined at design-time and are largely non-configurable.

*Supervisor Address Translation and Protection* (**satp**) is a special CSR (\acrshort{csr}):

> "This register holds the **physical page number (PPN)** of the root page table, i.e., its supervisor physical address divided by 4KiB;
> an **address space identifier (ASID)**, which facilitates address-translation fences on a per-address-space basis;
> and the **MODE field**, which selects the current address-translation scheme."
>
> `r tufte::quote_footer('--- The RISC-V Instruction Set Manual - Volume II: Privileged Architecture\ [@rv5vol2, section 4.1.11], emphasis: us.')`

(ref:cap-03-satp) Layout of the **satp** register.

```{r 03-satp, fig.cap='(ref:cap-03-satp)', }
knitr::include_graphics(normalizePath('fig/satp.pdf'))
```

Of these, **satp** is the only register that has any effect in Gotos and is active as long as the processor is in user-mode and the MODE field is set\ [@rv5vol2, page 75].
Gotos currently does not provide physical memory protection and does not allow setting physical memory attributes.

The PPN field of **satp** serves as a pointer to the page-table that is currently in use, aligned to a page-boundary.
Left-shifting the value by 12 bits will give the address of the first page in a multi-level page-table.

The ASID field is useful to prevent excessive flushing of TLBs as entries will be tagged with the current ASID.
Later an `sfence.vma` instruction can specify which ASID it should apply to or even which mapping.

#### The Sv32 page-table format {#arch:sv32}

The Sv32 format is the only format supported on 32-bit RISC-V processors\ [@rv5vol2, section 4.3].
It is a two-level page-table format that divides virtual addresses into three parts: **vpn1** (10b), **vpn0** (10b), and **page offset** (12b) (see figure\ \@ref(fig:03-sv32addr)).
The *virtual page number* (VPN) is split in two (**vpn1** and **vpn0**) as each part is used to index into a different level in the page table.

```{r 03-sv32addr, fig.cap='Segments of a virtual address in the Sv32 format.', out.width="50%"}
knitr::include_graphics(normalizePath('fig/sv32addr.pdf'))
```

A "level" in this page-table format always takes up a full page and is aligned on a page/frame boundary.
Page-table entries (PTEs) are 32 bits and pages/frames are 4096 bits, meaning a level contains 1024 entries, which fits with the 10 bits per level of the **vpnX**.

```{r 03-sv32pte, fig.cap='A RISC-V Sv32 PTE. The first 22 bits are the physical frame number; the last 10 bits are for different flags.', out.width="60%" }
knitr::include_graphics(normalizePath('fig/pte.pdf'))
```

A PTE works so that if the *R*, *X*, and *W* bits are all 0, and the *V* bit is 1, this entry is a pointer to the next level of the page-table\ (see table 4.5 of the privileged specification [@rv5vol2, page 80]).

Translation starts by using the **PPN** field of the **satp** register as a pointer to the first level of the page-table.
A detailed explanation of RISC-V address translation is found in the specification\ [@rv5vol2, section 4.3.2].
A brief summary is provided here:

(1) Let $p$ be the value of the **PPN** field in the **satp** register, and let $v_1$ be the value of **vpn1**.
    The current PTE is located at $4096p + 4v_1$.

(1) If the *V* bit of this PTE is set and $R + V + W \neq 0$ (at least one of the permission bits are set), this is a super-page; let $f$ be the value of the **PFN** field, $v_0$ be the value of **vpn0**, and $o$ be the **offset**, then the final address is given as $4096f + 4096v_0 + o$.
    Otherwise, move on to the next step.
    (Note that super-pages have to be aligned on a super-page boundary of 4MiB).

(1) Let $p$ be the value of $f$.
    The current PTE is located at $4096p + 4v_0$.

(1) Let $f$ be the value of the **PFN** field of the PTE, then the final address is given as $4096f + o$ (assuming the PTE is valid and sufficient permissions are present).

The steps outlined have missing pieces and make assumptions that we don't encounter invalid translations.
In these cases, appropriate page-faults are raised depending on the type of access that triggered the page-fault.

---

> **The Ex10 page-table format**
>
> A simpler way to explain this format is by using a smaller, but similar two-level page-table format that we will call Ex10 (example 10-bit).
> In this format we split 10-bit virtual addresses into three pieces: **vpn1** (3b), **vpn0** (3b), and an **offset** (4b).
> Thus, pages/frames are 16 bytes each (by the 4-bit offset).
> We use a PTE-size of 2 bytes, meaning each page/frame can hold 8 entries, corresponding to the 3 bits of each part of the **vpnX**.
>
> An Ex10 PTE consists of an 8-bit **PFN** and 8 flag bits.
>
> By using a relatively small format, we can do the translation visually as shown in figure\ \@ref(fig:03-ex10).


```{r 03-ex10, fig.cap='An example using the Ex10 page-table format to translate a 10-bit virtual address.', out.width='5in'}
knitr::include_graphics(normalizePath('fig/ex10.pdf'))
```

> Here we have found the value of `v1` by following the address step by step.
> The first level of the page-table is stored in the frame that starts at address `20`.
> We calculate the position of the first PTE as `24` (row 2, column 4).
> We take the **PFN** value of `04` to the next step.
> We repeat this one more time and we find that `v1` is located at `33` (row 3, column 3) and has a value of `AB`.
>
> Notice that while virtual addresses are only 10 bits, the resulting physical address can be as large as 12 bits (8-bit frame number and 4-bit offset).
> This is the case for Sv32 as well which translates 32-bit virtual addresses to 34-bit physical addresses, meaning a full 16GiB can be in use, even though processes can only map 4GiB.

---

### The program counter/instruction pointer

The program counter usually advances by 4 bytes after each cycle, pointing to the next instruction.
The exceptions are for jumps caused by control-flow instructions such as branches and jumps, and for traps (which are jumps to system-code).

### Traps {#arch:trap}

Much like in a real system, a trap is the only mechanism to switch from user-mode applications into OS code.
RISC-V specifies multiple reasons that a processor might trap and how traps should be implemented on a processor.
Gotos implements most of these trap reasons.
Some are ignored as they only occur in special modes or simply never occur because of the way Gotos does certain things.

Traps may be caused for several reasons:

- an illegal instruction may have been fetched,
- page fault because of insufficient permissions,
- misaligned access,
- timer interrupt,
- external interrupt, or
- an `ecall` or `ebreak` instruction was executed.

A few other trap reasons exist for certain modes or actions that are not available in Gotos.
The specification informs what should happen when a trap occurs:

1. The privilege level is elevated to machine mode (unless delegated, but we do not allow this in Gotos).

1. Virtualisation mode is set to 0 and information about the trap reason and the originating privilege mode is written to specific registers.
   In Gotos, the originating privilege mode will always be user mode so these registers are ignored.
   The trap cause is stored to **mcause**, the program counter is written to **mepc** (machine exception program counter), and the faulting value (if any) is written to **mtval**.

1. After trap setup is complete, the program counter is set to the value contained in **mtvec**, potentially with some offset for vectored exceptions.
   On this point, Gotos is very different.
   Gotos ignores the **mtvec** register, opting instead for a `HandleTrap(...)` function.
   This function is the main entry point for the operating system kernel to execute any code.

1. When trap handling is complete, a special instruction -- `mret` -- should be executed to set the program counter to the value stored in **mepc** and restore the privilege mode that was written in the second step.
   This means that, unless anything else is written to **mepc**, execution will continue at the address where the processor originally trapped.
   This is the intended behaviour, as many traps require OS intervention before retrying the instruction.
   Since the privilege mode is always user-mode, the only thing `mret` does is set the program counter to the value stored in **mepc**.

Gotos makes this last step implicit as it should be the only mechanism by which the OS changes the program counter.
The `trap` function used in the emulator is shown in listing\ \@ref(lst:03-trap).

The first two lines are part of the trap setup; the third line ensures that the processor will not automatically update the program counter after finishing; the fourth line calls the actual `HandleTrap(c)`, more about this in the next chapter; and the last line restores the program counter.

(ref:cap-03-trap) The `trap` function used in Gotos.

```{.go #lst:03-trap caption="(ref:cap-03-trap)"}
func (c *Core) trap(reason uint32) {
    c.csr[Csr_MCAUSE] = reason
    c.csr[Csr_MEPC] = c.pc
    c.jumped = true
    c.system.HandleTrap(c)
    c.pc = c.csr[Csr_MEPC]
}
```

Listing\ \@ref(lst:03-yield) shows a practical example of a trap-handler which returns control to the program that caused it.
It accomplishes this by fetching the original address, then writing the address of the next instruction to ensure that execution continues as normal.

(ref:cap-03-breakpoint) The simplest possible handler for a trap caused by a process executing `ebreak`.

```{.go #lst:03-yield caption="(ref:cap-03-breakpoint)"}
func (s *System) handleBreakpoint(c *cpu.Core) {
    trapAddress := c.GetCSR(cpu.Csr_MEPC)
    c.SetCSRc(cpu.Csr_MEPC, trapAddress+4)
}
```

The `ecall`-instruction (environment call) will usually be the method of choice for intentionally causing a trap; e.g. when performing a system call.
Here is a short piece of assembly which, when assembled and run on the emulator, will call the `exit` system call (provided the implementation follow guidelines put forth in this document):

```{.asm}
li    a0, 1  # load system call number
ecall        # cause trap
```
